### 2024-06-11 paper/tex/03-declarative-system.tex
corrections

\paragraph{Quantifiers}  
Symmetric rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} specify 
subtyping between top-level quantified types.
Usually, the polymorphic subtyping is represented by two rules
introducing quantifiers to the left and to the right-hand side of subtyping.
For conciseness of representation, we compose these rules into one.
First, our rule extends context $[[Γ]]$ with the quantified variables 
from the right-hand side ($[[pbs]]$ or $[[nbs]]$), 
as these variables must remain abstract.
Second, it verifies that the left-hand side quantifiers
($[[pas]]$ or $[[nas]]$) can be instantiated to continue subtyping recursively. 
This is the point of non-determinism as the instantiation can be chosen arbitrarily

@@

\paragraph{Invariant Shifts}
As mentioned above, an important restriction that we put on the subtyping system
is that subtyping of shifted types requires their equivalence, as shown in
\ruleref{\ottdruleDOneShiftDLabel} and \ruleref{\ottdruleDOneShiftULabel}. The
reason for this is that if both of these rules were relaxed to the covariant
form, the subtyping relation would become equivalent to the standard subtyping
of \systemf, which is undecidable \cite{tiuryn-urzczyn-96}. However, it is
suggested after certain changes \ruleref{\ottdruleDOneShiftULabel} can be
relaxed to the covariant form, thereby increasing the expressiveness of the
system. These changes are discussed in \cref{sec:weakening-invariant}.

\paragraph{Functions}
Standardly, subtyping of function types is covariant in the return type
and contravariant in the argument type.

\paragraph{Variables}
Subtyping of variables is defined reflexively, which is enough to ensure the
reflexivity of subtyping in general. The algorithm---specifically the least
upper bound procedure---will use the fact that the subtypes of a variable
coincide with its supertypes (\cref{prop:var-no-subtypes}),
which however does not hold for arbitrary types.

@@

\paragraph{Negative greatest lower bound might not exist}
The symmetric construction---the greatest lower bound of two negative types---does 
not always exist, as the following counterexample shows.
Consider the following types: 
\begin{itemize}
  \item $[[iN]]$ and $[[iQ]]$ are arbitrary closed types, 
  \item $[[iP]]$, $[[iP1]]$, and $[[iP2]]$ are non-equivalent closed types 
    such that $[[· ⊢ iP1 ≥ iP]]$ and $[[· ⊢ iP2 ≥ iP]]$, and 
    none of the types is equivalent to $[[iQ]]$.
\end{itemize}
What is the greatest common subtype of  
$[[iQ → ↓↑iQ → ↓↑iQ → iN]]$ and $[[iP → ↓↑iP1 → ↓↑iP2 → iN]]$?
The type $[[∀α⁺,β⁺,γ⁺. α⁺ → ↓↑β⁺ → ↓↑γ⁺ → iN]]$ is a common subtype,
however, it is not the greatest one, as it is too general.

@@

\begin{figure}[h]
  \ottdefnsDTLabeled
  \caption{Declarative Inference}
  \label{fig:declarative-inference}
\end{figure}

@@

\paragraph{Variables}
  Rule \ruleref{\ottdruleDTVarLabel} allows us to infer
  the type of a variable from the context. 
  In literature, one can find another version of this rule,
  that enables inferring a type \emph{equivalent}
  to the type from the context. 
  In our case, the inference of equivalent types
  is admissible in general case by \ruleref{\ottdruleDTPEquivLabel}.

\paragraph{Annotations}
  The annotation rules \ruleref{\ottdruleDTNAnnotLabel}
  and \ruleref{\ottdruleDTPAnnotLabel} use subtyping.
  The annotation is only valid if the
  inferred type is a subtype of the annotation type.

\paragraph{Abstractions}
  The typing of lambda abstraction is standard. 
  Rule \ruleref{\ottdruleDTtLamLabel} first checks
  that the given type annotating the argument is well-formed,
  and then infers the type of the body in the extended context.
  As a result, it returns an arrow type of function from the
  annotated type of the argument to the type of the body.
  Rule \ruleref{\ottdruleDTTLamLabel} infers polymorphic $[[∀]]$-type. 
  It extends the type context with the quantifying variable $[[pa]]$ and 
  infers the type of the body. As a result, it returns a polymorphic type
  quantifying the abstracted variable $[[pa]]$ over the type of the body.

\paragraph{Return and thunk}
  Rules \ruleref{\ottdruleDTReturnLabel} 
  and \ruleref{\ottdruleDTThunkLabel}
  simply add the corresponding shift constructors to the type of the body.

\paragraph{Unpack}
  Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
  First, it infers the normalized type of the existential package.
  The normalization is required to fix the order of the quantifying variables
  to bind them. Alternative approaches that do not require normalization
  will be discussed in \cref{sec:explicit-type-application}.
  After the bind, the rule infers the type of the body 
  and checks that it does not use the bound variables so that they do not 
  escape the scope.

\paragraph{Let binders}
  Rule \ruleref{\ottdruleDTVarLetLabel} is a standard rule for typing let
  binders: we infer the type of the bound value and continue the typing of the
  computation in the extended context.

@@

\paragraph{Applicative let binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.

@@

\paragraph{Typing up to equivalence}
  As discussed in \cref{sec:decl-equivalence}, the subtyping, as a preorder, 
  induces a non-trivial equivalence relation on types. 
  The system must not distinguish between equivalent types,
  and thus, type inference must be defined up to equivalence. 
  For this purpose, we use rules \ruleref{\ottdruleDTPEquivLabel}  
  and \ruleref{\ottdruleDTNEquivLabel}.
  They allow one to replace the inferred type with an equivalent one.

\paragraph{Application to an empty list of arguments}
  The base case of the application type inference is 
  represented by rule \ruleref{\ottdruleDTEmptyAppLabel}.
  If the head of the type $[[iN]]$ is applied to no arguments, 
  the type of the result is allowed to be $[[iN]]$ or any 
  equivalent type. We need to relax this rule up to equivalence
  to ensure the corresponding property globally:
  the inferred application type can be replaced with an equivalent one.
  Alternatively, we could have added a separate rule similar to 
  \ruleref{\ottdruleDTPEquivLabel}, however, 
  the local relaxation is sufficient to prove the global property.

\paragraph{Application of a polymorphic type $[[∀]]$}
  The complexity of the system lives in the rules whose output type is not
  immediately defined by their input and the output of their premises (\aka not
  mode-correct \cite{dunfield2020:bidirectional}). In our typing system,
  \ruleref{\ottdruleDTForallAppLabel} is such a rule: the instantiation of the
  quantifying variables is not known a priori. The algorithm we present in
  \cref{sec:algorithm} delays this instantiation until more information about it
  (in particular, typing constraints) is collected.

@@

\paragraph{Application of an arrow type}
  Another application rule is \ruleref{\ottdruleDTArrowAppLabel}.
  This is where the subtyping is used to check that the type of the argument
  is convertible to (a subtype of) the type of the function parameter.
  In the algorithm (\cref{sec:algorithm}), this subtyping check will provide the constraints
  we need to resolve the delayed instantiations of the quantifying variables.

@@

\paragraph{Type-level Translation}

@@

\paragraph{Term-level Translation}

>>>>

\paragraph*{Quantifiers}  
Symmetric rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} specify 
subtyping between top-level quantified types.
Usually, the polymorphic subtyping is represented by two rules
introducing quantifiers to the left and to the right-hand side of subtyping.
For conciseness of representation, we compose these rules into one.
First, our rule extends context $[[Γ]]$ with the quantified variables 
from the right-hand side ($[[pbs]]$ or $[[nbs]]$), 
as these variables must remain abstract.
Second, it verifies that the left-hand side quantifiers
($[[pas]]$ or $[[nas]]$) can be instantiated to continue subtyping recursively. 
This is the point of non-determinism as the instantiation can be chosen arbitrarily

@@

\paragraph*{Invariant Shifts}
As mentioned above, an important restriction that we put on the subtyping system
is that subtyping of shifted types requires their equivalence, as shown in
\ruleref{\ottdruleDOneShiftDLabel} and \ruleref{\ottdruleDOneShiftULabel}. The
reason for this is that if both of these rules were relaxed to the covariant
form, the subtyping relation would become equivalent to the standard subtyping
of \systemf, which is undecidable \cite{tiuryn-urzczyn-96}. However, it is
suggested after certain changes \ruleref{\ottdruleDOneShiftULabel} can be
relaxed to the covariant form, thereby increasing the expressiveness of the
system. These changes are discussed in \cref{sec:weakening-invariant}.

\paragraph*{Functions}
Standardly, subtyping of function types is covariant in the return type
and contravariant in the argument type.

\paragraph*{Variables}
Subtyping of variables is defined reflexively, which is enough to ensure the
reflexivity of subtyping in general. The algorithm---specifically the least
upper bound procedure---will use the fact that the subtypes of a variable
coincide with its supertypes (\cref{prop:var-no-subtypes}),
which however does not hold for arbitrary types.

@@

\paragraph*{Negative greatest lower bound might not exist}
The symmetric construction---the greatest lower bound of two negative types---does 
not always exist, as the following counterexample shows.
Consider the following types: 
\begin{itemize}
  \item $[[iN]]$ and $[[iQ]]$ are arbitrary closed types, 
  \item $[[iP]]$, $[[iP1]]$, and $[[iP2]]$ are non-equivalent closed types 
    such that $[[· ⊢ iP1 ≥ iP]]$ and $[[· ⊢ iP2 ≥ iP]]$, and 
    none of the types is equivalent to $[[iQ]]$.
\end{itemize}
What is the greatest common subtype of  
$[[iQ → ↓↑iQ → ↓↑iQ → iN]]$ and $[[iP → ↓↑iP1 → ↓↑iP2 → iN]]$?
The type $[[∀α⁺,β⁺,γ⁺. α⁺ → ↓↑β⁺ → ↓↑γ⁺ → iN]]$ is a common subtype,
however, it is not the greatest one, as it is too general.

@@

\begin{figure}[t!]
  \ottdefnsDTLabeled
  \caption{Declarative Inference}
  \label{fig:declarative-inference}
\end{figure}

@@

\paragraph*{Variables}
  Rule \ruleref{\ottdruleDTVarLabel} allows us to infer
  the type of a variable from the context. 
  In literature, one can find another version of this rule,
  that enables inferring a type \emph{equivalent}
  to the type from the context. 
  In our case, the inference of equivalent types
  is admissible in general case by \ruleref{\ottdruleDTPEquivLabel}.

\paragraph*{Annotations}
  The annotation rules \ruleref{\ottdruleDTNAnnotLabel}
  and \ruleref{\ottdruleDTPAnnotLabel} use subtyping.
  The annotation is only valid if the
  inferred type is a subtype of the annotation type.

\paragraph*{Abstractions}
  The typing of lambda abstraction is standard. 
  Rule \ruleref{\ottdruleDTtLamLabel} first checks
  that the given type annotating the argument is well-formed,
  and then infers the type of the body in the extended context.
  As a result, it returns an arrow type of function from the
  annotated type of the argument to the type of the body.
  Rule \ruleref{\ottdruleDTTLamLabel} infers polymorphic $[[∀]]$-type. 
  It extends the type context with the quantifying variable $[[pa]]$ and 
  infers the type of the body. As a result, it returns a polymorphic type
  quantifying the abstracted variable $[[pa]]$ over the type of the body.

\paragraph*{Return and thunk}
  Rules \ruleref{\ottdruleDTReturnLabel} 
  and \ruleref{\ottdruleDTThunkLabel}
  simply add the corresponding shift constructors to the type of the body.

\paragraph*{Unpack}
  Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
  First, it infers the normalized type of the existential package.
  The normalization is required to fix the order of the quantifying variables
  to bind them. Alternative approaches that do not require normalization
  will be discussed in \cref{sec:explicit-type-application}.
  After the bind, the rule infers the type of the body 
  and checks that it does not use the bound variables so that they do not 
  escape the scope.

\paragraph*{Let binders}
  Rule \ruleref{\ottdruleDTVarLetLabel} is a standard rule for typing let
  binders: we infer the type of the bound value and continue the typing of the
  computation in the extended context.

@@

\paragraph*{Applicative let binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.

@@

\paragraph*{Typing up to equivalence}
  As discussed in \cref{sec:decl-equivalence}, the subtyping, as a preorder, 
  induces a non-trivial equivalence relation on types. 
  The system must not distinguish between equivalent types,
  and thus, type inference must be defined up to equivalence. 
  For this purpose, we use rules \ruleref{\ottdruleDTPEquivLabel}  
  and \ruleref{\ottdruleDTNEquivLabel}.
  They allow one to replace the inferred type with an equivalent one.

\paragraph*{Application to an empty list of arguments}
  The base case of the application type inference is 
  represented by rule \ruleref{\ottdruleDTEmptyAppLabel}.
  If the head of the type $[[iN]]$ is applied to no arguments, 
  the type of the result is allowed to be $[[iN]]$ or any 
  equivalent type. We need to relax this rule up to equivalence
  to ensure the corresponding property globally:
  the inferred application type can be replaced with an equivalent one.
  Alternatively, we could have added a separate rule similar to 
  \ruleref{\ottdruleDTPEquivLabel}, however, 
  the local relaxation is sufficient to prove the global property.

\paragraph*{Application of a polymorphic type $[[∀]]$}
  The complexity of the system lives in the rules whose output type is not
  immediately defined by their input and the output of their premises (\aka not
  mode-correct \cite{dunfield2020:bidirectional}). In our typing system,
  \ruleref{\ottdruleDTForallAppLabel} is such a rule: the instantiation of the
  quantifying variables is not known a priori. The algorithm we present in
  \cref{sec:algorithm} delays this instantiation until more information about it
  (in particular, typing constraints) is collected.

@@

\paragraph*{Application of an arrow type}
  Another application rule is \ruleref{\ottdruleDTArrowAppLabel}.
  This is where the subtyping is used to check that the type of the argument
  is convertible to (a subtype of) the type of the function parameter.
  In the algorithm (\cref{sec:algorithm}), this subtyping check will provide the constraints
  we need to resolve the delayed instantiations of the quantifying variables.

@@

\paragraph*{Type-level Translation}

@@

\paragraph*{Term-level Translation}
